\vspace{-1.5em} \bf Towards Pandemic-Scale Ancestral Recombination Graphs of SARS-CoV-2

### 2023-06-06 paper.tex
Minor updates

\subsection{False positive recombinants}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to large numbers of samples, and are almost certainly
artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
in the Long ARG that are likely false positives, and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869),
labelled A--D. These false positives are associated with the origins
of large outbreaks. For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC). Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of 3; Section~\ref{sec:ls}).
The column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

@@

A varying degree of missingness is present in the genetic data, that is, for a
given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
As a result, we had to consider missingness in our implementation of the LS
Viterbi algorithm. One approach would be to `impute' all missing data
beforehand using external software to hard-call or define a distribution over
the possible missing nucleotides, and use that in the determination of path
probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide, but
rather to consider such sites as uninformative to the path probability, and
impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
the emission probability from unknown to
to any other state as 1.
% $\mbox{N} \rightarrow \star$ as 1.

@@

Given a newly attached node, mutation collapsing inspects its siblings from
previous sample days to check if any of them share (a subset of) the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations, and make that new node the parent of the
siblings carrying those shared mutations (Figure \ref{fig:overview_sc2ts}D). The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse.
Table~\ref{tab:args} shows that
while these nodes constitute a small fraction of the
overall total (Wide ARG: 2.77\%, Long ARG: 4.82\%),
they are parental to a much larger fraction of the
edges (Wide ARG: 12.77\%, Long ARG: 20.47\%).

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent which carries all its mutations except those
causing the reversions on the newly added node (Figure \ref{fig:overview_sc2ts}E).
Table~\ref{tab:args} shows that
these nodes constitute a similar fraction of the
overall numbers of nodes (Wide ARG: 2.79\%, Long ARG: 4.39\%)
and edges (Wide ARG: 12.66\%, Long ARG: 18.46\%)
as mutation collapsing.

Together these operations substantially increase the overall parsimony
of the inferred ARGs, most noticeably in terms of the number of
reversion mutations inferred. These nodes are also being chosen by the
LS HMM as likely choices of parent (as indicated by the node and edge
counts quoted above), also demonstrating that the heuristics are successfully
capturing features of real sequences.
However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Because the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual, it is clear that there
is room for improvement and that further parsimony-increasing
heuristics
(e.g., resolving reversions beyond those on immediately adjacent edges)
would likely be of benefit .

>>>>

\subsection{False positive recombinants}
The Long and Wide ARGs both contain several recombination nodes that are
ancestral to large numbers of samples, and are almost certainly
artefactual. Table~\ref{tab:false_positive} shows details for four recombinants
in the Long ARG that are likely false positives, and have come to be
the ancestors of a large number of samples. These are the top-four recombinants
from the Long ARG in terms of numbers of descendant samples
(the fifth-largest has substantially fewer, at 14,869 descendant samples),
labelled A--D. These false positives are associated with the origins
of large outbreaks. For each row, we show the sample's Pango lineage
designation and the rank of that sample in the overall list of
samples with that designation. Thus, rows C and D are the first
samples seen in the Long ARG from the BA.2.9 and BA.3 lineages,
and A and B are the third and eighth samples from the B.1.617.2
lineage (Delta VoC), respectively. Also shown is the overall ``cost'' of the Viterbi
solution computed by the LS HMM (Section~\ref{sec:ls}), which
is $3 \times$ number of recombinations + number of additional mutations
(for a mismatch ratio of 3; Section~\ref{sec:ls}).
The column shows that each of these samples was a large
``distance'' from the current ARG when it was added. The mean
HMM cost over all 2,078 recombinants in the Long ARG is 8.39
(median: 7), and the mean cost for the  763 non-singleton
recombinants is 8.5 (median: 7). Thus, the recombinants in
Table~\ref{tab:false_positive} are in the tail of this distribution.

@@

A varying degree of missingness is present in the genetic data, that is, for a
given sample, a subset of nucleotides may be coded as `unknown' at a set of positions.
As a result, we had to consider missingness in our implementation of the LS
Viterbi algorithm. One approach would be to `impute' all missing data
beforehand using external software to hard-call or define a distribution over
the possible missing nucleotides, and use that in the determination of path
probabilities. Another, which we carry out within \texttt{sc2ts}, is to not attempt to estimate the true nucleotide, but
rather to consider such sites as uninformative to the path probability, and
impute missing data within that path on the fly, placing them within the emission probability transition from unknown to
the emission probability from unknown to
to any other state as 1.

@@

Given a newly attached node, mutation collapsing inspects its siblings from
previous sample days to check if any of them share (a subset of) the mutations that
it carries. If so, we increase the overall parsimony of the inference
by creating a new node representing the ancestor that carried
those shared mutations, and make that new node the parent of the
siblings carrying those shared mutations (Figure \ref{fig:overview_sc2ts}D). The patterns of
shared mutations between siblings can be complex, and the current
implementation uses a simple greedy strategy for choosing
the particular mutations to collapse.

The reversion push operation inspects a newly added node to see
if any of its mutations are ``immediate reversions''; that is,
are reversions of a mutation that occurred on the new node's
immediate parent. We increase the overall parsimony of the
inference by ``pushing in'' a new node which descends from the
original parent, and carries all its mutations except those
causing the reversions on the newly added node (Figure \ref{fig:overview_sc2ts}E).

Table~\ref{tab:args} shows that nodes generated by 
these operations constitute 
roughly the same fraction of the total in both the Long and Wide ARGs,
and contribute significantly to the overall topology.
These nodes are also being chosen by the
LS HMM as likely choices of parent (data not shown) 
demonstrating that the heuristics are successfully
capturing features of real sequences.
However, they are both simple
greedy operations, just examining the local parts of the
ARG topology affected by newly added samples. Because the inferred
ARGs still contain a large number of reversion mutations which
are likely to be mostly artefactual (Section~\ref{sec:mutation_spectrum}), 
it is clear that there
is room for improvement and that further parsimony-increasing
heuristics
(e.g., resolving reversions beyond those on immediately adjacent edges)
would likely be of benefit .
